---
title: "Biophysical model of estimated cfDNA tumor fraction for colorectal cancer"
author: "J. Bredno"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output:
  html_document:
    fig.width: 12
---

This markdown creates tables, lists, and figures to create and validate the
biophysical model of colorectal cancer.

```{r setup, include = FALSE}
# Basic setup. Setup markdown properties, source dependent functions,
# confirm availability of required packages
requireNamespace("knitr")
requireNamespace("tidyverse")
`%>%` <- magrittr::`%>%`
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(digits = 4)
source("../R/biophysical_modeling.R")
```

```{r load_data_for_modeling, include = FALSE}
df_crc_in <- readr::read_tsv("../data/supplemental_data_crc_training.tsv")
df_crc <- df_crc_in
```

### Figure 10c

WGBS classifier score by ECTF for colorectal cancer. Samples are colored by clinical
stage and samples with imputed ECTF are shown as triangles.


```{r impute_tumor_fraction}
# For simpler re-use, imputed ECTF is already part of the shared data.
# The previously imputed data is removed and re-created here for complete
# reproducibility
set.seed(0)
df_crc <- df_crc %>%
  dplyr::mutate(ectf_check = ectf,
                ectf = ifelse(ectf_source == "imputed",
                              NA, ectf)) %>%
  dplyr::select(-ectf_source)

lret <- impute_ectf(df_crc,
                    replace_with_imputed = TRUE,
                    draw_or_fixed = "draw",
                    score_frac_min = 0.02,
                    score_frac_max = 0.98)
df_crc <- lret$df_model
print(lret$hPlot %>%
        assign_figure_style(axis_x = "ECTF",
                            axis_y = "WGBS classifier score",
                            color_legend = "Clinical stage",
                            shape_legend = "Tumor fraction source",
                            legend_position = c(0.8, 0.4),
                            color_points = TRUE,
                            export = "Fig_10c_crc_impute_ectf"))
```

```{r derive_depth_of_invasion}
# nolint start
lookup_microinvasion <- tibble::tribble(
  ~microinvasion, ~quant_invasion, ~display_microinvasion, ~score_invasion,
  "Other/missing", -1, "Other/missing", NA,
  "Not stated", -1,  "Other/missing", NA,
  "Lamina propria/muscularis mucosae (intramucosal carcinoma)", 1, "Intramucosal", 0,
  "Submucosa", 2, "Submucosa", 0,
  "Muscularis propria", 3, "Muscularis propria", 0,
  "Into the subserosal adipose tissue or the nonperitonealized pericolic or perirectal soft tissues",
  4, "Subserosa", 1,
  "Penetrates visceral peritoneum (serosa)", 5, "Penetrates serosa", 1,
  "Tumor penetrates to the surface of the visceral peritoneum (serosa) and directly invades adjacent structures",
  5, "Penetrates serosa", 1,
  "Invades adjacent structures", 6, "Invades adjacent", 1)
# nolint end
df_crc <- df_crc %>%
  dplyr::left_join(lookup_microinvasion, by = "microinvasion") %>%
  dplyr::mutate(quant_invasion = as.integer(quant_invasion),
                imputed_invasion = dplyr::case_when(
                  !is.na(score_invasion) ~ score_invasion,
                  t_stage %in% c("T0", "T1", "T2") ~ 0,
                  t_stage %in% c("T3", "T4") ~ 1,
                  TRUE ~ NA_real_),
                display_microinvasion = factor(display_microinvasion,
                                                  c("Other/missing",
                                                    "Intramucosal",
                                                    "Submucosa",
                                                    "Muscularis propria",
                                                    "Subserosa",
                                                    "Penetrates serosa",
                                                    "Invades adjacent")))
```

```{r derive_size_ln_assay}
set.seed(0)
l_ret <- derive_primary_size(df_crc,
                             l_imputation = NULL,
                             draw_or_fixed = "draw",
                             dimension = 2)
df_crc <- l_ret$df_model
l_imputation <- l_ret$l_imputation

df_crc <- df_crc %>%
  get_ln_status() %>%
  assay_quant()
```

### Figures 10A and B

Estimated cfDNA tumor fraction is shown for all invasive cancers (not stage 0) with
known clinical stage in the stage breakdown and for all cases with reported T-stage
in the T-stage breakdown

```{r display_tf}
print(basic_violin(df_crc %>%
                     dplyr::filter(clinical_stage %in% c("I", "II", "III", "IV")),
                   "clinical_stage", "ectf",
                   shape = "ectf_source") %>%
        assign_figure_style(log_y = TRUE,
                            axis_x = "Clinical stage",
                            axis_y = "ECTF",
                            shape_legend = "Tumor fraction source",
                            legend_position = c(0.15, 0.85),
                            export = "Fig_10a_crc_stage_ectf"))
print(basic_violin(df_crc %>%
                     dplyr::filter(quant_invasion != -1),
                   "display_microinvasion", "ectf",
                   shape = "ectf_source") %>%
        assign_figure_style(log_y = TRUE,
                            axis_x = "Depth of microinvasion",
                            axis_y = "ECTF",
                            x_axis_vertical = TRUE,
                            shape_legend = "Tumor fraction source",
                            legend_position = c(0.15, 0.85),
                            export = "Fig_10b_crc_depth_ectf"))
```

### Table 3

Analytical Model for Colorectal Cancer.

```{r modeling}
df_crc <- df_crc %>%
  dplyr::mutate(area_shallow = ifelse(imputed_invasion == 0, primary_area, 0),
                area_deep = ifelse(imputed_invasion == 1, primary_area, 0),
                score_histograde = dplyr::case_when(
                  histologic_grade == "1" ~ -1,
                  histologic_grade == "2" ~ 0,
                  histologic_grade == "3" ~ 1,
                  histologic_grade == "4" ~ 1,
                  TRUE ~ 0))

v_variables <- c("area_shallow", "area_deep", "ln_status", "score_histograde")
v_targets <- c("ectf", "ctdna_amount_ge")

modeling <- create_model(df_crc %>%
                           dplyr::filter(clinical_stage %in% c("I", "II", "III")),
                         v_targets = v_targets,
                         v_variables = v_variables,
                         v_constraints = c(TRUE, TRUE, TRUE, TRUE))
```

```{r display_analytical_model}
knitr::kable(modeling$df_analysis_model)
summary(modeling$analysis_model)
```

Non-negativity constraints affected `r modeling$violate_constraints`.

```{r create_prediction_and_quant_models}
predict_deep <- lm(data = df_crc %>%
                     dplyr::filter(clinical_stage %in% c("I", "II", "III"),
                                   score_invasion == 1),
                   ectf ~ 0 + area_deep)

predict_shallow <- lm(data = df_crc %>%
                        dplyr::filter(clinical_stage %in% c("I", "II", "III"),
                                      score_invasion == 0),
                      ectf ~ 0 + area_shallow)

quantify_deep <- lm(data = df_crc %>%
                      dplyr::filter(clinical_stage %in% c("I", "II", "III"),
                                    score_invasion == 1),
                    ctdna_amount_ge ~ 0 + area_deep)

quantify_shallow <- lm(data = df_crc %>%
                         dplyr::filter(clinical_stage %in% c("I", "II", "III"),
                                       score_invasion == 0),
                       ctdna_amount_ge ~ 0 + area_shallow)
```


```{r save models for command line prediction tool}
saveRDS(predict_deep, file = "../cmd/crc_deep_prediction_model.Rds")
saveRDS(predict_shallow, file = "../cmd/crc_shallow_prediction_model.Rds")
saveRDS(quantify_deep, file = "../cmd/crc_deep_quantitative_model.Rds")
saveRDS(quantify_shallow, file = "../cmd/crc_shallow_quantitative_model.Rds")
saveRDS(l_imputation, file = "../cmd/crc_impute_size.Rds")
```


Quantitative model for deep invasion.

```{r summary_predict_deep}
summary(quantify_deep)
```

Quantitative model for shallow invasion.

```{r summary_quantify_shallow}
summary(quantify_shallow)
```

### Figure 11

```{r visualize prediction model}
df_plot <- df_crc %>%
  dplyr::filter(clinical_stage %in% c("I", "II", "III")) %>%
  dplyr::mutate(invasion = ifelse(score_invasion == 1, "deep", "shallow"))

hPlot <- ggplot2::ggplot() +
  ggplot2::geom_point(data = df_plot %>% dplyr::filter(score_invasion == 1),
                      ggplot2::aes(x = area_deep, y = ectf,
                                   color = clinical_stage, shape = invasion)) +
  ggplot2::geom_smooth(data = df_plot %>% dplyr::filter(score_invasion == 1),
                       ggplot2::aes(x = area_deep, y = ectf), color = "black",
                       method = lm, formula = y ~ x) +
  ggplot2::geom_point(data = df_plot %>% dplyr::filter(score_invasion == 0),
                      ggplot2::aes(x = area_shallow, y = ectf,
                                   color = clinical_stage, shape = invasion)) +
  ggplot2::geom_smooth(data = df_plot %>% dplyr::filter(score_invasion == 0),
                       ggplot2::aes(x = area_shallow, y = ectf), color = "red",
                       method = lm, formula = y ~ x)

assign_figure_style(hPlot,
                    axis_x = "Tumor surface area",
                    axis_y = "ECTF", log_x = TRUE, log_y = TRUE,
                    color_legend = "Clinical stage",
                    shape_legend = "Microinvasion",
                    color_points = TRUE,
                    export = "Fig_11_crc_model")
```

### Figure 3B (right)

These case numbers are used in the Consort diagrams in Figure 3.

```{r breakdown_training}
invasive_adenocarcinoma <-
  df_crc$clinical_stage != "0"
measured_ectf <-
  df_crc$ectf_source == "measured" & !is.na(df_crc$ectf)
imputed_ectf <-
  df_crc$ectf_source == "imputed" & !is.na(df_crc$ectf)
has_ectf <-
  !is.na(df_crc$ectf)
stage_i_ii_iii <-
  df_crc$clinical_stage %in% c("I", "II", "III")
has_size <-
  !is.na(df_crc$primary_area)
has_invasion <-
  !is.na(df_crc$imputed_invasion)
has_ln <-
  !is.na(df_crc$ln_status)

df_population <- tibble::tribble(
  ~population, ~n,
  "Invasive ADC of colon",
  sum(invasive_adenocarcinoma),
  "~ with measured ECTF",
  sum(invasive_adenocarcinoma & measured_ectf),
  "~ with imputed ECTF",
  sum(invasive_adenocarcinoma & imputed_ectf),
  "~ with any ECTF",
  sum(invasive_adenocarcinoma & has_ectf),
  "Stages I, II, III",
  sum(has_ectf & stage_i_ii_iii),
  "~ with size information",
  sum(has_ectf & stage_i_ii_iii & has_size),
  "~ with microinvasion information",
  sum(has_ectf & has_size & stage_i_ii_iii & has_invasion),
  "~ with LN information",
  sum(has_ectf & stage_i_ii_iii & has_ln & has_size & has_invasion),
  "Validation",
  sum(has_ectf & stage_i_ii_iii & has_ln & has_invasion & has_size),
  "~ Stage I",
  sum(has_ectf & has_ln & has_invasion & has_size &
          df_crc$clinical_stage == "I"),
  "~ Stage II",
  sum(has_ectf & has_ln & has_invasion & has_size &
          df_crc$clinical_stage == "II"),
  "~ Stage III",
  sum(has_ectf & has_ln & has_invasion & has_size &
          df_crc$clinical_stage == "III"))
knitr::kable(df_population)
```

### Training data imputation status

```{r report_training_imputation}
df_analytical_model <-
  df_crc[has_ectf & stage_i_ii_iii & has_ln & has_invasion & has_size, ]
l_report_imputation <- report_imputation(df_analytical_model)
data.frame(data_source = stringr::str_sub(names(l_report_imputation), 3),
           n_samples = as.integer(l_report_imputation)) %>%
  knitr::kable()
```

```{r read validation data}
df_validation <- readr::read_tsv("../data/supplemental_data_crc_validation.tsv")
```

```{r repeat_derivations}
set.seed(0)
l_ret <- df_validation %>%
  derive_primary_size(l_imputation = l_imputation,
                      draw_or_fixed = "draw",
                      dimension = 2)
df_validate_crc <- l_ret$df_model %>%
  get_ln_status() %>%
  dplyr::left_join(lookup_microinvasion, by = "microinvasion") %>%
  dplyr::mutate(imputed_invasion = dplyr::case_when(
    !is.na(score_invasion) ~ score_invasion,
    t_stage %in% c("T0", "T1", "T2") ~ 0,
    t_stage %in% c("T3", "T4") ~ 1,
    TRUE ~ NA_real_))
```

```{r model_predictions}
predict_crc <- function(df_data) {
  v_predict_deep <- predict(predict_deep, df_data)
  v_predict_shallow <- predict(predict_shallow, df_data)
  df_data$ectf_predicted <- NA
  is_shallow <- df_data$imputed_invasion == 0 & !is.na(df_data$imputed_invasion)
  is_deep <- df_data$imputed_invasion == 1 & !is.na(df_data$imputed_invasion)
  df_data$ectf_predicted[is_deep] <- v_predict_deep[is_deep]
  df_data$ectf_predicted[is_shallow] <- v_predict_shallow[is_shallow]
  df_data
}

df_validate_crc <- df_validate_crc %>%
  dplyr::mutate(area_shallow = ifelse(score_invasion == 0, primary_area, 0),
                area_deep = ifelse(score_invasion == 1, primary_area, 0))
df_validate_crc <- predict_crc(df_validate_crc)
v_correction <- df_validate_crc$ectf_predicted
```

### Figure 12A and B

```{r validation result plots}
df_results <- df_validate_crc %>%
  dplyr::filter(clinical_stage %in% c("I", "II", "III"),
                !is.na(ectf_predicted)) %>%
  dplyr::mutate(ectf_visualize = pmin(1.0, ectf_predicted))

hPlot <- waterfall_plot(df_results %>%
                          dplyr::mutate(detected = (tm_classifier_result == "detected")),
                        name_y = "ectf_visualize",
                        color_1 = "detected",
                        color_2 = "clinical_stage") +
  ggplot2::scale_fill_manual(labels = c("I", "II", "III",
                                        "not detected", "detected"),
                             values = c("black", "gray", "white",
                                        "yellow", "blue")) +
  ggplot2::ylab("ECTF")
print(hPlot)
ggplot2::ggsave(filename = "../figures/Fig_12a_crc_waterfall.eps",
                plot = hPlot,
                dpi = 600, device = "eps")
ggplot2::ggsave(filename = "../figures/Fig_12a_crc_waterfall.svg",
                plot = hPlot,
                dpi = 600, device = "svg")

l_ret <- simple_roc(df_results %>%
                      dplyr::mutate(detected = (tm_classifier_result == "detected")),
                    "detected", "ectf_predicted")

print(l_ret$hPlot %>%
        assign_figure_style(axis_x = "False positive rate",
                                axis_y = "True positive rate",
                                export = "Fig_12b_crc_roc"))
```

Model-predicted ECTF explains detection with TM assay with an AUC of `r l_ret$auc`
(95% CI `r as.numeric(l_ret$auc_ci[c(1,3)])`).
Detected cases have higher model-predicted ECTF than undetected cases with one-sided two-sample Wilcoxon Rank Sum test p-Value `r l_ret$wilcoxon_p`.

```{r validation_participants}
df_validate_crc %>%
  dplyr::count(clinical_stage, tm_classifier_result) %>%
  tidyr::pivot_wider(names_from = "tm_classifier_result",
                     values_from = "n", values_fill = 0) %>%
  dplyr::mutate(n = detected + `not detected`) %>%
  dplyr::select(clinical_stage, n, detected) %>%
  knitr::kable()

df_validate_crc %>%
  dplyr::filter(!is.na(area_deep)) %>%
  dplyr::count(clinical_stage, tm_classifier_result) %>%
  tidyr::pivot_wider(names_from = "tm_classifier_result",
                     values_from = "n", values_fill = 0) %>%
  dplyr::mutate(n = detected + `not detected`) %>%
  dplyr::select(clinical_stage, n, detected) %>%
  knitr::kable()
```

### Figure 3c (right)

These case numbers are used in the Consort diagrams in Figure 3.

```{r breakdown_validation_population}
invasive_adenocarcinoma <-
  df_validate_crc$clinical_stage != "0"
stage_i_ii_iii <-
  df_validate_crc$clinical_stage %in% c("I", "II", "III")
has_size <-
  !is.na(df_validate_crc$primary_area)
has_invasion <-
  !is.na(df_validate_crc$imputed_invasion)
has_ln <-
  !is.na(df_validate_crc$ln_status)

df_population <- tibble::tribble(
  ~population, ~n,
  "Invasive ADC of colon",
  sum(invasive_adenocarcinoma),
  "Stages I, II, III",
  sum(stage_i_ii_iii),
  "~ with size information",
  sum(stage_i_ii_iii & has_size),
  "~ with microinvasion information",
  sum(has_size & stage_i_ii_iii & has_invasion),
  "~ with LN information",
  sum(stage_i_ii_iii & has_ln & has_size & has_invasion),
  "~ Stage I",
  sum(has_ln & has_invasion & has_size &
          df_validate_crc$clinical_stage == "I"),
  "~ Stage II",
  sum(has_ln & has_invasion & has_size &
          df_validate_crc$clinical_stage == "II"),
  "~ Stage III",
  sum(has_ln & has_invasion & has_size &
          df_validate_crc$clinical_stage == "III"))
knitr::kable(df_population)
```

### Validation data imputation status

```{r report_validation_imputation}
df_used_in_validation <-
  df_validate_crc[stage_i_ii_iii & has_invasion & has_size, ]
l_report_imputation <- report_imputation(df_used_in_validation)
data.frame(data_source = stringr::str_sub(names(l_report_imputation), 3),
           n_samples = as.integer(l_report_imputation),
           stringsAsFactors = FALSE) %>%
  # ectf is always model-predicted for validation
  dplyr::filter(!endsWith(data_source, "ectf")) %>%
  knitr::kable()
```

### Supplemental figure 3: ROC broken down by stage

```{r roc_by_stage}
lret <- combined_roc(df_results %>%
                       dplyr::mutate(detected = (tm_classifier_result == "detected")),
                     "detected", "ectf_predicted", "clinical_stage")
print(lret$hPlot %>%
        assign_figure_style(axis_x = "False positive rate",
                            axis_y = "True positive rate",
                            color_legend = "Clinical stage",
                            legend_position = c(0.8, 0.3),
                            color_points = TRUE,
                            export = "Fig_Sup3_crc_ROC"))
knitr::kable(tibble::tribble(
  ~stage, ~auc, ~ci, ~wilcoxon_p,
  "I", as.numeric(lret$auc_I), lret$auc_ci_I, lret$wilcoxon_p_I,
  "II", as.numeric(lret$auc_II), lret$auc_ci_II, lret$wilcoxon_p_II,
  "III", as.numeric(lret$auc_III), lret$auc_ci_III,  lret$wilcoxon_p_III,
  "all", as.numeric(lret$auc_all), lret$auc_ci_all,  lret$wilcoxon_p_all)
)
```

### Reproducibility information

```{r reproducibility_info}
sessionInfo()
```
