---
title: "Biophysical model of circulating tumor fraction for lung cancer"
author: "J. Bredno"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output:
  html_document:
    fig.width: 12
---

This markdown creates tables, lists, and figures to create and validate the
biophysical model of lung cancer.

```{r setup, include = FALSE}
# Basic setup. Setup markdown properties, source dependent functions,
# confirm availability of required packages
requireNamespace("knitr")
requireNamespace("tidyverse")
`%>%` <- magrittr::`%>%`
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(digits = 4)
source("../R/biophysical_modeling.R")
```

```{r load_data_for_modeling, include = FALSE}
df_luc_in <- readr::read_tsv("../data/supplemental_data_luc_training.tsv")
df_luc <- df_luc_in
```

### Figure 7B

WGBS classifier score by cTF for lung cancer. Samples are colored by clinical
stage and samples with imputed cTF are shown as triangles.

```{r impute_tumor_fraction}
# For simpler re-use, imputed cTF is already part of the shared data.
# The previously imputed data is removed and re-created here for complete
# reproducibility
set.seed(0)
df_luc <- df_luc %>%
  dplyr::mutate(ctf_check = ctf,
                ctf = ifelse(ctf_source == "imputed",
                              NA, ctf)) %>%
  dplyr::select(-ctf_source)

lret <- impute_ctf(df_luc,
                    replace_with_imputed = TRUE,
                    draw_or_fixed = "draw",
                    score_frac_min = 0.02,
                    score_frac_max = 0.98)
df_luc <- lret$df_model
print(lret$hPlot %>%
        assign_figure_style(axis_x = "cTF",
                            axis_y = "WGBS classifier score",
                            color_legend = "Clinical stage",
                            shape_legend = "Tumor fraction source",
                            legend_position = c(0.8, 0.4),
                            color_points = TRUE,
                            export = "Fig_7b_luc_impute_ctf"))
```

```{r derive_size_ln_assay}
set.seed(0)
l_ret <- derive_primary_size(df_luc,
                             l_imputation = NULL,
                             draw_or_fixed = "draw",
                             dimension = 3)
df_luc <- l_ret$df_model
l_imputation <- l_ret$l_imputation

df_luc <- df_luc %>%
  get_ln_status() %>%
  assay_quant()
```

### Figure 7A

Estimated cfDNA tumor fraction by clinical stage for lung cancer is shown for
all invasive cancers (not stage 0) with known clinical stage.

```{r tf_by_clinical}
print(basic_violin(df_data = df_luc %>%
                     dplyr::filter(clinical_stage %in% c("I", "II", "III", "IV")),
                   name_x = "clinical_stage",
                   name_y = "ctf",
                   shape = "ctf_source") %>%
        assign_figure_style(log_y = TRUE,
                            axis_x = "Clinical stage",
                            axis_y = "cTF",
                            shape_legend = "Tumor fraction source",
                            legend_position = c(0.15, 0.85),
                            export = "Fig_7a_luc_stage_ctf"))
```

### Table 2

Analytical Model for Lung Cancer.

```{r modeling}
df_luc <- df_luc %>%
  dplyr::mutate(elg = primary_volume * (fdg_suv - 1.0),
                score_histograde = dplyr::case_when(
                  histologic_grade == "1" ~ -1,
                  histologic_grade == "2" ~ 0,
                  histologic_grade == "3" ~ 1,
                  TRUE ~ 0),
                is_adc = as.numeric(histologic_type == "ADC"),
                is_scc = as.numeric(histologic_type == "SCC"),
                is_sclc = as.numeric(histologic_type == "SCLC"))
v_variables <- c("elg", "ln_status", "score_histograde", "is_adc", "is_scc", "is_sclc")
v_targets <- c("ctf", "ctdna_amount_ge")

modeling <- create_model(df_luc %>%
                           dplyr::filter(clinical_stage %in% c("I", "II", "III")),
                         v_targets = v_targets,
                         v_variables = v_variables,
                         v_constraints = c(TRUE, TRUE, TRUE, FALSE, FALSE, FALSE))
```

```{r save models for command line prediction tool}
saveRDS(modeling$prediction_model, file = "../cmd/luc_prediction_model.Rds")
saveRDS(modeling$quantitative_model, file = "../cmd/luc_quantitative_model.Rds")
saveRDS(l_imputation, file = "../cmd/luc_impute_size.Rds")
```

```{r modeling_result}
knitr::kable(modeling$df_analysis_model %>%
               dplyr::select(-p_value) %>%
               dplyr::mutate(variable = dplyr::case_when(
                               variable == "elg" ~ "ELG",
                               variable == "ln_status" ~ "Lymph node status",
                               variable == "score_histograde" ~ "Histologic grade",
                               variable == "is_adc" ~ "Adenocarcinoma",
                               variable == "is_scc" ~ "Squamous cell carcinoma",
                               variable == "is_sclc" ~ "Small cell lung cancer",
                               TRUE ~ variable)))
```

Non-negativity constraints affected (can be empty):
`r modeling$violate_constraints`

Quantitative model for Lung Cancer.

```{r quantitative_model}
summary(modeling$quantitative_model)
```

### Figure 8

```{r visualize prediction model}
hPlot <- ggplot2::ggplot(data = df_luc %>%
                           dplyr::filter(clinical_stage %in%
                                           c("I", "II", "III"))) +
  ggplot2::geom_point(mapping = ggplot2::aes(x = elg,
                                             y = ctf,
                                             color = clinical_stage)) +
  ggplot2::geom_smooth(mapping = ggplot2::aes(x = elg, y = ctf),
                       method = lm,
                       formula = y ~ x)
assign_figure_style(hPlot,
                    axis_x = "ELG", axis_y = "cTF", log_x = TRUE, log_y = TRUE,
                    color_legend = "Clinical stage",
                    legend_position = c(0.85, 0.2),
                    color_points = TRUE,
                    export = "Fig_8_luc_model")
```

### Figure 3B (center)

These case numbers are used in the Consort diagrams in Figure 3.

```{r training_population}
invasive_luc <-
  df_luc$clinical_stage != "0"
measured_ctf <-
  df_luc$ctf_source == "measured" & !is.na(df_luc$ctf)
imputed_ctf <-
  df_luc$ctf_source == "imputed" & !is.na(df_luc$ctf)
has_ctf <-
  !is.na(df_luc$ctf)
stage_i_ii_iii <-
  df_luc$clinical_stage %in% c("I", "II", "III")
has_size <-
  !is.na(df_luc$primary_volume)
has_suv <-
  !is.na(df_luc$fdg_suv)
has_ln <-
  !is.na(df_luc$ln_status)

df_population <- tibble::tribble(
  ~population, ~n,
  "Invasive lung cancer",
  sum(invasive_luc),
  "~ with measured cTF",
  sum(invasive_luc & measured_ctf),
  "~ with imputed cTF",
  sum(invasive_luc & imputed_ctf),
  "~ with any cTF",
  sum(invasive_luc & has_ctf),
  "Stages I, II, III",
  sum(has_ctf & stage_i_ii_iii),
  "~ with size information",
  sum(has_ctf & stage_i_ii_iii & has_size),
  "~ with PET SUV information",
  sum(has_ctf & has_size & stage_i_ii_iii & has_suv),
  "~ with LN information",
  sum(has_ctf & stage_i_ii_iii & has_ln & has_size & has_suv),
  "Validation",
  sum(has_ctf & stage_i_ii_iii & has_ln & has_suv & has_size),
  "~ Stage I",
  sum(has_ctf & has_ln & has_suv & has_size &
          df_luc$clinical_stage == "I"),
  "~ Stage II",
  sum(has_ctf & has_ln & has_suv & has_size &
          df_luc$clinical_stage == "II"),
  "~ Stage III",
  sum(has_ctf & has_ln & has_suv & has_size &
          df_luc$clinical_stage == "III"))
knitr::kable(df_population)
```

### Training data imputation status

```{r report_imputation}
df_analytical_model <-
  df_luc[has_ctf & stage_i_ii_iii & has_ln & has_suv & has_size, ]
l_report_imputation <- report_imputation(df_analytical_model)
data.frame(data_source = stringr::str_sub(names(l_report_imputation), 3),
           n_samples = as.integer(l_report_imputation)) %>%
  knitr::kable()
```

```{r read validation data}
df_validation <- readr::read_tsv("../data/supplemental_data_luc_validation.tsv")
```

```{r repeat_derivations}
set.seed(0)
l_ret <- df_validation %>%
  derive_primary_size(l_imputation = l_imputation,
                      draw_or_fixed = "draw",
                      dimension = 3)
df_validate_luc <- l_ret$df_model %>%
  get_ln_status()
```

```{r model_predictions}
df_validate_luc <- df_validate_luc %>%
  dplyr::mutate(elg = primary_volume * (fdg_suv - 1.0))

df_validate_luc$ctf_predicted <- predict(modeling$prediction_model,
                                         df_validate_luc)
```

### Figure 9A and B

```{r validation result plots}
df_results <- df_validate_luc %>%
  dplyr::filter(clinical_stage %in% c("I", "II", "III"),
                !is.na(ctf_predicted)) %>%
  dplyr::mutate(ctf_visualize = pmin(1.0, ctf_predicted))

hPlot <- waterfall_plot(df_results %>%
                          dplyr::mutate(detected = (tm_classifier_result == "detected")),
                        name_y = "ctf_visualize",
                        color_1 = "detected",
                        color_2 = "clinical_stage") +
  ggplot2::scale_fill_manual(labels = c("I", "II", "III",
                                        "not detected", "detected"),
                             values = c("black", "gray", "white",
                                        "yellow", "blue")) +
  ggplot2::ylab("cTF")
print(hPlot)
ggplot2::ggsave(filename = "../figures/Fig_9a_luc_waterfall.eps",
                plot = hPlot,
                dpi = 600, device = "eps")
ggplot2::ggsave(filename = "../figures/Fig_9a_luc_waterfall.svg",
                plot = hPlot,
                dpi = 600, device = "svg")

l_ret <- simple_roc(df_results %>%
                      dplyr::mutate(detected = (tm_classifier_result == "detected")),
                    "detected", "ctf_predicted")
print(l_ret$hPlot %>%
        assign_figure_style(axis_x = "False positive rate",
                            axis_y = "True positive rate",
                            export = "Fig_9b_luc_roc"))
```

Model-predicted cTF explains detection with TM assay with an AUC of `r l_ret$auc`
(95% CI `r as.numeric(l_ret$auc_ci[c(1,3)])`).
Detected cases have higher model-predicted cTF than undetected cases with one-sided two-sample Wilcoxon Rank Sum test p-Value `r l_ret$wilcoxon_p`.

### Figure 3C (center)

These case numbers are used in the Consort diagrams in Figure 3.

```{r consort_validation}
invasive_luc <-
  df_validate_luc$clinical_stage != "0"
stage_i_ii_iii <-
  df_validate_luc$clinical_stage %in% c("I", "II", "III")
has_size <-
  !is.na(df_validate_luc$primary_volume)
has_suv <-
  !is.na(df_validate_luc$fdg_suv)
has_ln <-
  !is.na(df_validate_luc$ln_status)

df_population <- tibble::tribble(
  ~population, ~n,
  "Invasive Lung Cancer",
  sum(invasive_luc),
  "Stages I, II, III",
  sum(stage_i_ii_iii),
  "~ with size information",
  sum(stage_i_ii_iii & has_size),
  "~ with PET SUV information",
  sum(has_size & stage_i_ii_iii & has_suv),
  "~ with LN information",
  sum(stage_i_ii_iii & has_ln & has_size & has_suv),
  "~ Stage I",
  sum(has_ln & has_suv & has_size &
          df_validate_luc$clinical_stage == "I"),
  "~ Stage II",
  sum(has_ln & has_suv & has_size &
          df_validate_luc$clinical_stage == "II"),
  "~ Stage III",
  sum(has_ln & has_suv & has_size &
          df_validate_luc$clinical_stage == "III"))
knitr::kable(df_population)
```

### Detection with TM assay for validation participants

```{r validation_participants}
df_validate_luc %>%
  dplyr::filter(!is.na(elg),
                !is.na(ln_status),
                clinical_stage %in% c("I", "II", "III")) %>%
  dplyr::count(clinical_stage, tm_classifier_result) %>%
  tidyr::pivot_wider(names_from = "tm_classifier_result",
                     values_from = "n", values_fill = 0) %>%
  dplyr::mutate(n = detected + `not detected`) %>%
  dplyr::select(clinical_stage, n, detected) %>%
  knitr::kable()
```

### Validation data imputation status

```{r report_validation_imputation}
df_used_in_validation <-
  df_validate_luc[stage_i_ii_iii & has_suv & has_size, ]
l_report_imputation <- report_imputation(df_used_in_validation)
data.frame(data_source = stringr::str_sub(names(l_report_imputation), 3),
           n_samples = as.integer(l_report_imputation),
           stringsAsFactors = FALSE) %>%
  # cTF is always model-predicted for validation
  dplyr::filter(!endsWith(data_source, "ctf")) %>%
  knitr::kable()
```

### Supplemental table 2: Competitive prediction model

```{r competing_model}
v_variables <- c("elg", "primary_volume", "ln_status", "score_histograde",
                 "is_adc", "is_scc", "is_sclc")
v_targets <- c("ctf", "ctdna_amount_ge")
competing_model <- create_model(df_luc %>%
                                      dplyr::filter(clinical_stage %in% c("I", "II", "III")),
                                    v_targets = v_targets, v_variables = v_variables)
knitr::kable(competing_model$df_analysis_model %>%
               dplyr::select(-p_value) %>%
               dplyr::mutate(variable = dplyr::case_when(
                               variable == "elg" ~ "ELG",
                               variable == "ln_status" ~ "Lymph node status",
                               variable == "score_histograde" ~ "Histologic grade",
                               variable == "is_adc" ~ "Adenocarcinoma",
                               variable == "is_scc" ~ "Squamous cell carcinoma",
                               variable == "is_sclc" ~ "Small cell lung cancer",
                               variable == "primary_volume" ~ "Primary volume",
                               TRUE ~ variable)))

v_variables <- c("primary_volume", "ln_status", "score_histograde",
                 "is_adc", "is_scc", "is_sclc")
competing_model <- create_model(df_luc %>%
                                      dplyr::filter(clinical_stage %in% c("I", "II", "III")),
                                    v_targets = v_targets, v_variables = v_variables)

knitr::kable(competing_model$df_analysis_model %>%
               dplyr::select(-p_value) %>%
               dplyr::mutate(variable = dplyr::case_when(
                               variable == "elg" ~ "ELG",
                               variable == "ln_status" ~ "Lymph node status",
                               variable == "score_histograde" ~ "Histologic grade",
                               variable == "is_adc" ~ "Adenocarcinoma",
                               variable == "is_scc" ~ "Squamous cell carcinoma",
                               variable == "is_sclc" ~ "Small cell lung cancer",
                               variable == "primary_volume" ~ "Primary volume",
                               TRUE ~ variable)))
```

### Supplemental figure 2: ROC broken down by stage

```{r roc_by_stage}
lret <- combined_roc(df_results %>%
                       dplyr::mutate(detected = (tm_classifier_result == "detected")),
                     "detected", "ctf_predicted", "clinical_stage")
print(lret$hPlot %>%
        assign_figure_style(axis_x = "False positive rate",
                            axis_y = "True positive rate",
                            color_legend = "Clinical stage",
                            legend_position = c(0.8, 0.3),
                            color_points = TRUE,
                            export = "Fig_Sup3_crc_ROC"))
knitr::kable(tibble::tribble(
  ~stage, ~auc, ~ci, ~wilcoxon_p,
  "I", as.numeric(lret$auc_I), lret$auc_ci_I, lret$wilcoxon_p_I,
  "II", as.numeric(lret$auc_II), lret$auc_ci_II, lret$wilcoxon_p_II,
  "III", as.numeric(lret$auc_III), lret$auc_ci_III,  lret$wilcoxon_p_III,
  "all", as.numeric(lret$auc_all), lret$auc_ci_all,  lret$wilcoxon_p_all)
)
```

### Reproducibility information

```{r reproducibility_info}
sessionInfo()
```
